---
title: "Psoariasis Severity - Data Integration III"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Microbiome

## raw counts

```{r}
#| label: raw 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

library(ggplot2)
library(phyloseq)
library(dplyr)
library(tidyverse)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")
sample_info <- read.table("../../data/sample_md.tsv", header = T, sep = "\t", check.names = F)
row.names(sample_info) <- sample_info$SampleID

sample_info$Type <- factor(sample_info$Type, levels = c("control", "pso_nonlesional", "pso_lesional"))
sample_info$Gender <- as.factor(sample_info$Gender)

sample_info <- sample_info %>% mutate(PsoTreatment = replace(PsoTreatment, PsoTreatment == "0", "None"))
sample_info$PsoTreatment <- factor(sample_info$PsoTreatment, levels = c("None", "Moisturizer", "TopicalSteroids"))

sample_info$PASISeverity <- as.factor(sample_info$PASISeverity)
sample_info <- sample_info %>% 
  mutate(PASISeverity = case_when(
    PASISeverity == "0" ~ "None",
    PASISeverity == "1" ~ "Low",
    PASISeverity == "2" ~ "Medium",
    PASISeverity == "3" ~ "High",
    TRUE ~ PASISeverity))
sample_info$PASISeverity <- factor(sample_info$PASISeverity, levels = c("None", "Low", "Medium", "High"))

sample_info$TypeSeverity <- factor(sample_info$TypeSeverity, 
  levels = c("control_0", "pso_nonlesional_1", "pso_nonlesional_2", "pso_nonlesional_3",
  "pso_lesional_1", "pso_lesional_2", "pso_lesional_3"))

sample_info <- sample_info %>% mutate(PrevSysTreatmentPso = replace(PrevSysTreatmentPso, PrevSysTreatmentPso == "", "None"))
sample_info$PrevSysTreatmentPso <- factor(sample_info$PrevSysTreatmentPso, levels = c("None", "No", "Yes"))

sample_info$AntiinflamMedication <- as.factor(sample_info$AntiinflamMedication)
sample_info$SSRI <- as.factor(sample_info$SSRI)
sample_info$OtherMedication <- as.factor(sample_info$OtherMedication)
sample_info$DiabetesHypertensionHyperlipidemia <- as.factor(sample_info$DiabetesHypertensionHyperlipidemia)

sample_info <- sample_info %>% 
  mutate(OtherDiseases = case_when(
    OtherDiseases == "" ~ NA,
    OtherDiseases == "0" ~ "None",
    TRUE ~ OtherDiseases))
sample_info$OtherDiseases <- as.factor(sample_info$OtherDiseases)

sample_info <- sample_info %>% mutate(SmokingSnuff = replace(SmokingSnuff, SmokingSnuff == "0", "None"))
sample_info$SmokingSnuff <- as.factor(sample_info$SmokingSnuff)

sample_info$Alcohol <- as.factor(sample_info$Alcohol)
sample_info <- sample_info %>% 
  mutate(Alcohol = case_when(
    Alcohol == "0" ~ "None",
    Alcohol == "1" ~ "Low",
    Alcohol == "2" ~ "Medium",
    Alcohol == "3" ~ "High",
    Alcohol == "4" ~ "VeryHigh",
    TRUE ~ Alcohol))
sample_info$Alcohol <- factor(sample_info$Alcohol, levels = c("None", "Low", "Medium", "High", "VeryHigh"))

sample_info <- sample_info %>% 
  mutate(Arthritis = case_when(
    Arthritis == "" ~ NA,
    Arthritis == "0" ~ "None",
    TRUE ~ Arthritis))
sample_info$Arthritis <- as.factor(sample_info$Arthritis)

sample_info <- sample_info %>% mutate(Location = replace(Location, Location == "", NA))
sample_info$Location <- as.factor(sample_info$Location)

lulu_otu_table <- read.table("../lulu_otu_table.tsv", header = T, row.names = 1, sep = "\t")

taxa_info <- read.table("../processed_taxa.tsv", header = 1, row.names = 1, sep = "\t")
taxa_info <- as.matrix(taxa_info)

OTU = otu_table(lulu_otu_table, taxa_are_rows = TRUE)
TAX = tax_table(taxa_info)
META = sample_data(sample_info)

psor_psq_raw = phyloseq(OTU, TAX, META)
psor_psq_raw
sum(otu_table(psor_psq_raw))
```

## Filtering based on taxa and counts

```{r}
#| label: filt 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

main_taxa <- c("Bacteria", "Archaea")
psor_psq_tFilt <- subset_taxa(psor_psq_raw, Kingdom %in% main_taxa)
psor_psq_tcFilt <- filter_taxa(psor_psq_tFilt, function(x) sum(x > 5) > (0.1*length(x)), TRUE)
psor_psq_filt <- subset_taxa(psor_psq_tcFilt, Family != "Mitochondria")

psor_psq_filt
sum(otu_table(psor_psq_filt))
```

More than **75%** of the counts are still kept in the dataset after the filteration steps below:
 - Only ASVs that are **Bacteria** or **Archaea** kept and the other ASVs are thrown out assuming that they are sequencing or PCR artefacts
 - ASVs that are present in atleast **10** samples with atleast **5 reads** are kept and the rest are filtered.
 - ASVs that were predicted to be **Mitochondria** are removed.

## Including --add_spe information from dada2

```{r}
#| label: addspe 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

add_spe <- read.table("../ampliseq_16S_output/dada2/ASV_tax_species.silva_138_2.tsv", header = 1, sep = "\t")
spe_exact <- add_spe %>%
  select(ASV_ID, Genus, Species_exact, confidence) %>%
  filter(Species_exact != "") %>%
  filter(confidence > 0.75) %>%
  mutate(Species_exact = paste0(Genus, '_', Species_exact)) %>%
  select(ASV_ID, Species_exact)

taxa_new <- tax_table(psor_psq_filt) %>% 
  as.data.frame() %>%
  rownames_to_column('ASV_ID') %>%
  left_join(spe_exact) %>%
  mutate(Species = ifelse(is.na(Species_exact),
    Species, Species_exact)) %>%
  select(-Species_exact)

row.names(taxa_new) <- taxa_new$ASV_ID
taxa_new$ASV_ID <- NULL
taxa_new <- as.matrix(taxa_new)
tax_table(psor_psq_filt) <- tax_table(taxa_new)

known_species <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Species)) 

known_species_asv_ids <- known_species %>%
  pull(ASV_ID)

known_genus <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Genus))

known_genus_asv_ids <- known_genus %>%
  pull(ASV_ID)

known_genus %>%
  pull(Genus) %>%
  n_distinct()

known_species %>%
  pull(Species) %>%
  n_distinct()
```

There are **367 known species** and **243 known genera** found in this dataset. 

## Genus and Species level tables

```{r}
#| label: gen-spe tables 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

gen_psq <- prune_taxa(psor_psq_filt, taxa = known_genus_asv_ids)
sum(otu_table(gen_psq))

spe_psq <- prune_taxa(psor_psq_filt, taxa = known_species_asv_ids)
sum(otu_table(spe_psq))

gen_glom_psq <- tax_glom(gen_psq, "Genus")
spe_glom_psq <- tax_glom(spe_psq, "Species")

```

## sort microbiome (leisional)

### check species to keep

```{r}
#| label: micr-sort
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

snp_filt <- read.table("../snp_filtered.tsv", row.names = 1, header = T)
low.count.removal <- function(
                        data, # OTU count df of size n (sample) x p (OTU)
                        percent=0.01 # cutoff chosen
                        ){
    keep.otu = which(colSums(data)*100/(sum(colSums(data))) > percent)
    data.filter = data[,keep.otu]
    return(list(data.filter = data.filter, keep.otu = keep.otu))
}

spe_lei <- subset_samples(spe_glom_psq, SampleID %in% row.names(snp_filt))
taxa_info <- as.data.frame(tax_table(spe_lei))
mic.raw <- as.data.frame(t(otu_table(spe_lei)))
names(mic.raw) <- taxa_info$Species
mic.offset <- mic.raw+1
result.filter <- low.count.removal(mic.offset, percent=0.01)
mic.filter <- result.filter$data.filter
length(result.filter$keep.otu) 
```

### check samples to keep

```{r}
#| label: micr-samp
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(mixOmics)
set.seed(100)

lib.size <- apply(mic.filter, 1, sum) # determine total count for each sample
barplot(lib.size) # and plot as bar plot
mic.filter <- mic.filter[which(lib.size < 200000 & lib.size > 15000),]
mic.clr <- logratio.transfo(as.matrix(mic.filter), 
                             logratio = 'CLR', offset = 0)
```

The outliers `K26MK` (very low library size) and `K36MK` (very high library size) are removed.

# SNP data 

## sort SNP 

```{r}
#| label: snp-sort
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

snp.filt <- snp_filt[which(row.names(snp_filt) %in% row.names(mic.clr)), ]
snp.filt <- snp.filt[order(row.names(snp.filt)), ]
mic.clr <- mic.clr[order(row.names(mic.clr)), ]

sample_info_ord <- snp.filt %>% 
  as.data.frame() %>%
  rownames_to_column('SampleID') %>%
  dplyr::select(SampleID) %>%
  left_join(sample_info) %>% 
  column_to_rownames(var = "SampleID")
```

# integration

```{r}
#| label: integ-I
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


library(mixOmics)
set.seed(100)

X <- list(mic = mic.clr,
          snp = snp.filt)

Y <- droplevels(sample_info_ord$Type)
summary(Y)

design <- matrix(0.1, ncol = length(X), nrow = length(X), 
                dimnames = list(names(X), names(X)))
diag(design) <- 0
design 
```

## Num. of components 

```{r}
#| label: integ-II
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

diablo.leisional <- block.plsda(X, Y, ncomp = 5, design = design)

set.seed(100) # For reproducibility, remove for your analyses
perf.diablo.leisional = perf(diablo.leisional, validation = 'Mfold', folds = 10, nrepeat = 10)

# Plot of the error rates based on weighted vote
plot(perf.diablo.leisional)
perf.diablo.leisional$choice.ncomp$WeightedVote
ncomp <- perf.diablo.leisional$choice.ncomp$WeightedVote["Overall.BER", "centroids.dist"]
```

## Num. of variables


```{r}
#| label: integ-III
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

set.seed(123) # for reproducibility
test.keepX <- list(mic = c(5:9, seq(10, 40, 5)),
                   snp = c(5:9, seq(10, 20, 2)))

# tune.diablo.leisional <- tune.block.splsda(X, Y, ncomp = 2, 
#                               test.keepX = test.keepX, design = design,
#                               validation = 'Mfold', folds = 10, nrepeat = 100, 
#                               BPPARAM = BiocParallel::SnowParam(workers = 12),
#                               dist = "centroids.dist")
# list.keepX <- tune.diablo.leisional$choice.keepX
list.keepX <- list(mic = c(20, 7), snp = c(16, 12))
```

## Final Model

### Samples

```{r}
#| label: integ-IV
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

diablo.leisional <- block.splsda(X, Y, ncomp = ncomp, 
                            keepX = list.keepX, design = design)
#plotDiablo(diablo.leisional, ncomp = 1)
plotIndiv(diablo.leisional, ind.names = FALSE, legend = TRUE, 
          title = 'Leisional, DIABLO comp 1 - 2', ellipse = T, abline=T)
```

```{r}
#| label: pub1
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/plotIndiv_leisional.svg", width = 10, height = 7)
plotIndiv(diablo.leisional, ind.names = FALSE, legend = TRUE, 
          ellipse = T, abline=T)
dev.off()
```

### Variables

```{r}
#| label: integ-V
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
#| fig-dpi: 300
#| fig-align: center

plotVar(diablo.leisional, var.names = FALSE, style = 'graphics', legend = TRUE, 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'brown1'),
        title = 'Leisional, DIABLO comp 1 - 2')

```

#### Circos Plot

```{r}
#| label: integ-VI
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


circosPlot(diablo.leisional, cutoff = 0.7, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1'),
           color.cor = c("chocolate3","grey20"), size.variables = 0.5)
```

```{r}
#| label: pub2
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/circosPlot_leisional.svg", width = 10, height = 10)
circosPlot(diablo.leisional, cutoff = 0.7, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1'),
           color.cor = c("chocolate3","grey20"), size.variables = 0.70, 
           size.labels = 1.5)
dev.off()
```

#### Loadings plot

```{r}
#| label: integ-VII
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

plotLoadings(diablo.leisional, comp = 1, contrib = 'max', method = 'median', size.name = 1.5)

# extracting microbiome loadings
mic_loadings_df <- data.frame(
  variable = rownames(diablo.leisional$loadings$mic),
  loading = diablo.leisional$loadings$mic[, 1],  # component 1
  comp = 1
)
mic_loadings_df <- mic_loadings_df %>%
                    filter(loading != 0) %>%
                    mutate(status = ifelse(loading > 0, "pso_leisional", "control"))
loadings_plot_mic <- ggplot(mic_loadings_df, aes(x = reorder(variable, abs(loading)), y = loading, fill = status)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "MIC",
    y = "Loadings value"
  ) +
  theme(axis.line.y = element_blank()) +
  scale_fill_manual(values = c("#567adf", "#df9138ff"))

# extracting snp loadings

snp_loadings_df <- data.frame(
  variable = rownames(diablo.leisional$loadings$snp),
  loading = diablo.leisional$loadings$snp[, 1],  # component 1
  comp = 1
)
snp_loadings_df <- snp_loadings_df %>%
                    filter(loading != 0) %>%
                    mutate(status = ifelse(loading > 0, "pso_leisional", "control"))
loadings_plot_snp <- ggplot(snp_loadings_df, aes(x = reorder(variable, abs(loading)), y = loading, fill = status)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "SNP",
    y = "Loadings value"
  ) +
  theme(axis.line.y = element_blank()) +
  scale_fill_manual(values = c("#567adf", "#df9138ff"))
```

```{r}
#| label: pub3
#| warning: false
#| message: false
#| eval: false


library(ggpubr)
svg("../Publication_figures/loadingsPlot_leisional.svg", width = 10, height = 8)
ggarrange(loadings_plot_mic, loadings_plot_snp, ncol = 2, common.legend=T, legend="right")
dev.off()
```

#### Heatmap

```{r}
#| label: integ-VIII
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


cim.res <- cimDiablo(diablo.leisional, transpose = T, color.blocks = c('darkorchid', 'brown1'),
          comp = 1, margin=c(8,20), legend.position = "topright")

mat <- cim.res$mat

df_long <- as.data.frame(mat) %>%
  rownames_to_column("Feature") %>%
  pivot_longer(-Feature, names_to = "Sample", values_to = "Value")

ggplot(df_long, aes(x = Sample, y = Feature, fill = Value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6)
  ) +
  labs(title = "Clustered Image Map (ggplot2 version)",
       fill = "Expression")
```

```{r}
#| label: pub4
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/heatMap_leisional.svg", width = 15, height = 10)
cimDiablo(diablo.leisional, transpose = T, color.blocks = c('darkorchid', 'brown1'),
          comp = 1, margin=c(8,20), legend.position = "topright",
          row.cex = 1.5, col.cex = 1.5, size.legend = 1)
dev.off()
```

## AUC & Prediction

```{r}
#| label: integ-IX
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


auc.diablo.leisional.mic <- auroc(diablo.leisional, roc.block = "mic", roc.comp = 2,
                   print = F)

auc.diablo.leisional.snp <- auroc(diablo.leisional, roc.block = "snp", roc.comp = 2,
                   print = F)

data.test.leision <- list(mic = mic.clr,
          snp = snp.filt)

predict.diablo.leision <- predict(diablo.leisional, newdata = data.test.leision)
confusion.mat.leision <- get.confusion_matrix(truth = droplevels(sample_info_ord$Type), 
                     predicted = predict.diablo.leision$WeightedVote$centroids.dist[,2])
confusion.mat.leision
```

# NMDS from latent

```{r}
#| label: nmds
#| warning: false
#| message: false
#| eval: false

cim.res <- cimDiablo(diablo.leisional, transpose = T, color.blocks = c('darkorchid', 'brown1'),
          comp = 1, margin=c(8,20), legend.position = "topright")

library(vegan)
set.seed(100)
lei_dist <- dist(as.data.frame(cim.res$mat))
lei_NMDS <- metaMDS(lei_dist, k=2)

lei_NMDS_long <- lei_NMDS$points %>%
  as.data.frame() %>%
  rownames_to_column('SampleID') %>%
  left_join(sample_info)

lei_nmds_plot <- ggplot(lei_NMDS_long) +
  geom_point(aes(MDS1, MDS2, color = Type), size = 4) +
  stat_ellipse(aes(MDS1, MDS2, color = Type), type = "t") +
  xlab('NMDS1') +
  ylab('NMDS2') +
  geom_hline(yintercept = 0, color = "lightgrey", linetype = 2) +
  geom_vline(xintercept = 0, color = "lightgrey", linetype = 2) +
  scale_color_manual(values = c("#567adf", "#df9138ff")) +
  theme_light(base_size = 16) +
  theme(panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank())

svg("../Publication_figures/nmds_leisional.svg", width = 10, height = 8)
lei_nmds_plot
dev.off()
```