---
title: "Psoariasis Severity - Data Integration V - Severity"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Microbiomes

The idea here is that since we have 2 microbiome for each patient with Psoriasis, we will treat these 2 microbes as 2 diffent omics from the same patient and integrate them with the SNP data. So, we will have 3 omics data that we can integrate to see how it correlates to severity.

```{r}
#| label: micr-sort
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(ggplot2)
library(phyloseq)
library(dplyr)
library(tidyverse)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")
snp_filt <- read.table("../snp_filtered.tsv", row.names = 1, header = T)

low.count.removal <- function(
                        data, # OTU count df of size n (sample) x p (OTU)
                        percent=0.01 # cutoff chosen
                        ){
    keep.otu = which(colSums(data)*100/(sum(colSums(data))) > percent)
    data.filter = data[,keep.otu]
    return(list(data.filter = data.filter, keep.otu = keep.otu))
}

#Leisional microbiome samples
spe_glom_psq <- readRDS("../spe_glom_psq.RDS")
spe_lei <- subset_samples(spe_glom_psq, Type == "pso_lesional")
taxa_info <- as.data.frame(tax_table(spe_lei))
mic.raw <- as.data.frame(t(otu_table(spe_lei)))
names(mic.raw) <- taxa_info$Species
mic.offset <- mic.raw+1
result.filter <- low.count.removal(mic.offset, percent=0.01)
mic.lei.filter <- result.filter$data.filter
length(result.filter$keep.otu) 

mic.lei.filter <- mic.lei.filter %>%
  rownames_to_column('SampleID') %>%
  mutate(SampleID = str_replace(SampleID, "(^M.+)P$", "\\1K")) %>%
  column_to_rownames(var = "SampleID")

#Non-Leisional microbiome samples
spe_non <- subset_samples(spe_glom_psq, Type == "pso_nonlesional")
taxa_info <- as.data.frame(tax_table(spe_non))
mic.raw <- as.data.frame(t(otu_table(spe_non)))
names(mic.raw) <- taxa_info$Species
mic.offset <- mic.raw+1
result.filter <- low.count.removal(mic.offset, percent=0.01)
mic.non.filter <- result.filter$data.filter
length(result.filter$keep.otu) 

```

## check samples to keep

### Leisional

```{r}
#| label: micr-samp
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(mixOmics)
set.seed(100)

lib.size <- apply(mic.lei.filter, 1, sum) # determine total count for each sample
barplot(lib.size) # and plot as bar plot
mic.lei.filter <- mic.lei.filter[which(lib.size > 20000),]
mic.lei.clr <- logratio.transfo(as.matrix(mic.lei.filter), 
                             logratio = 'CLR', offset = 0)
```

The outliers `M33MK` and `M35MK` (low library size) are removed.

### Non-Leisional

```{r}
#| label: micr-samp2
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


set.seed(100)

lib.size <- apply(mic.non.filter, 1, sum) # determine total count for each sample
barplot(lib.size) # and plot as bar plot
mic.non.filter <- mic.non.filter[which(lib.size < 250000),]
mic.non.clr <- logratio.transfo(as.matrix(mic.non.filter), 
                             logratio = 'CLR', offset = 0)
```

The outliers `M16MK` (high library size) is removed.

# SNP data 

