---
title: "Psoariasis Severity - SNP Markers"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Loading packages 

## raw data

```{r}
#| label: raw 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(ggplot2)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(vegan)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")
sample_info <- read.table("../../data/sample_md.tsv", header = T, sep = "\t", check.names = F)
row.names(sample_info) <- sample_info$SampleID

sample_info$Type <- factor(sample_info$Type, levels = c("control", "pso_nonlesional", "pso_lesional"))
sample_info$Gender <- as.factor(sample_info$Gender)

sample_info <- sample_info %>% mutate(PsoTreatment = replace(PsoTreatment, PsoTreatment == "0", "None"))
sample_info$PsoTreatment <- factor(sample_info$PsoTreatment, levels = c("None", "Moisturizer", "TopicalSteroids"))

sample_info$PASISeverity <- as.factor(sample_info$PASISeverity)
sample_info <- sample_info %>% 
  mutate(PASISeverity = case_when(
    PASISeverity == "0" ~ "None",
    PASISeverity == "1" ~ "Low",
    PASISeverity == "2" ~ "Medium",
    PASISeverity == "3" ~ "High",
    TRUE ~ PASISeverity))
sample_info$PASISeverity <- factor(sample_info$PASISeverity, levels = c("None", "Low", "Medium", "High"))

sample_info$TypeSeverity <- factor(sample_info$TypeSeverity, 
  levels = c("control_0", "pso_nonlesional_1", "pso_nonlesional_2", "pso_nonlesional_3",
  "pso_lesional_1", "pso_lesional_2", "pso_lesional_3"))

sample_info <- sample_info %>% mutate(PrevSysTreatmentPso = replace(PrevSysTreatmentPso, PrevSysTreatmentPso == "", "None"))
sample_info$PrevSysTreatmentPso <- factor(sample_info$PrevSysTreatmentPso, levels = c("None", "No", "Yes"))

sample_info$AntiinflamMedication <- as.factor(sample_info$AntiinflamMedication)
sample_info$SSRI <- as.factor(sample_info$SSRI)
sample_info$OtherMedication <- as.factor(sample_info$OtherMedication)
sample_info$DiabetesHypertensionHyperlipidemia <- as.factor(sample_info$DiabetesHypertensionHyperlipidemia)

sample_info <- sample_info %>% 
  mutate(OtherDiseases = case_when(
    OtherDiseases == "" ~ NA,
    OtherDiseases == "0" ~ "None",
    TRUE ~ OtherDiseases))
sample_info$OtherDiseases <- as.factor(sample_info$OtherDiseases)

sample_info <- sample_info %>% mutate(SmokingSnuff = replace(SmokingSnuff, SmokingSnuff == "0", "None"))
sample_info$SmokingSnuff <- as.factor(sample_info$SmokingSnuff)

sample_info$Alcohol <- as.factor(sample_info$Alcohol)
sample_info <- sample_info %>% 
  mutate(Alcohol = case_when(
    Alcohol == "0" ~ "None",
    Alcohol == "1" ~ "Low",
    Alcohol == "2" ~ "Medium",
    Alcohol == "3" ~ "High",
    Alcohol == "4" ~ "VeryHigh",
    TRUE ~ Alcohol))
sample_info$Alcohol <- factor(sample_info$Alcohol, levels = c("None", "Low", "Medium", "High", "VeryHigh"))

sample_info <- sample_info %>% 
  mutate(Arthritis = case_when(
    Arthritis == "" ~ NA,
    Arthritis == "0" ~ "None",
    TRUE ~ Arthritis))
sample_info$Arthritis <- as.factor(sample_info$Arthritis)

sample_info <- sample_info %>% mutate(Location = replace(Location, Location == "", NA))
sample_info$Location <- as.factor(sample_info$Location)

all_gen <- read.table("../../data/snp_raw.tsv", sep = "\t", header = T, check.names = F)
all_gen[all_gen == ''] <- NA
```

## Filter samples

Here we filter the SNP data to remove duplicates from the same patient and also the patients that have a lot of missing data. So, for each patient only the sample names that end `P` are kept. Then I also removed samples `'K23MK', 'M21MP' and 'M28MP'` because they have a lot of missing data (`23/49` markers have `NA` values).

```{r}
#| label: filt 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

target_samples <- sample_info %>% 
  filter(Type != "pso_nonlesional") %>%
  pull(SampleID)

all_gen_remDup <- all_gen %>%
  filter(SampleID %in% target_samples)

row.names(all_gen_remDup) <- all_gen_remDup$SampleID
all_gen_remDup$SampleID <- NULL
tail(sort(rowSums(is.na(all_gen_remDup))))

rem_samples <- c('K23MK', 'M21MP', 'M28MP')

all_gen_fin <- all_gen_remDup %>%
  rownames_to_column('SampleID') %>%
  filter(!(SampleID %in% rem_samples))

```

# Marker Statistics

```{r}
#| label: marker stats
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center



marker_data <- all_gen_fin %>% select(-SampleID)

# Calculate allele statistics for each SNP marker
allele_info <- apply(marker_data, 2, function(col) {
  alleles <- unlist(strsplit(na.omit(col), " "))  # Split genotypes into alleles
  if (length(alleles) == 0) return(NULL)  # Skip empty columns
  allele_freqs <- table(alleles) / length(alleles)  # Calculate allele frequencies
  list(
    NumAlleles = length(unique(alleles)),  # Count unique alleles
    MAF = if (length(allele_freqs) > 1) min(allele_freqs) else 0  # Calculate minor allele frequency (MAF)
  )
})

# Create a data frame with marker statistics (call rate, number of alleles, MAF)
marker_stats <- data.frame(
  Marker = colnames(marker_data),  # SNP marker names
  CallRate = colMeans(!is.na(marker_data)),  # Call rate
  NumAlleles = sapply(allele_info, function(info) if (!is.null(info)) info$NumAlleles else NA),  # Number of alleles
  MAF = sapply(allele_info, function(info) if (!is.null(info)) info$MAF else NA)  # MAF
)

# Preview the marker statistics
head(marker_stats)
```

## Filter Markers

```{r}
#| label: filter markers
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

valid_markers <- marker_stats %>%
  filter(CallRate >= 0.85,  # Call rate >= 85%
         NumAlleles <= 2,   # Bi-allelic markers
         MAF >= 0.05) %>%   # MAF >= 5%
  pull(Marker)  # Extract marker names

# Filter genotype data to include only valid markers and "CEL file ID"
filt_gen <- all_gen_fin %>%
  select(SampleID, all_of(valid_markers))  # Keep valid markers

# Check dimensions of filtered genotype data
dim(filt_gen)
```

## Convert to Numeric

```{r}
#| label: snp to ortho
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

convert_to_numeric <- function(geno_column) {
  geno_column <- as.character(geno_column)  # Ensure character format to handle factor variables correctly
  alleles <- names(sort(table(unlist(strsplit(na.omit(geno_column), " "))), decreasing = TRUE))  # Extract unique alleles by splitting the genotype string
  if (length(alleles) != 2) return(rep(NA, length(geno_column)))  # Skip non-biallelic SNPs by returning NA for all individuals
  ref_allele <- alleles[1]  # Define the baseline allele, e.g. reference allele
  alt_allele <- alleles[2]  # Define the other allele, e.g. alternate allele
  # Define a function to recode each genotype based on the defined alleles
  recode <- function(gt) {
    if (is.na(gt)) return(NA) # Handle missing values
    if (gt == paste0(ref_allele, " ", ref_allele)) return(0) # Homozygous (allele 1) = 0
    if (gt == paste0(ref_allele, " ", alt_allele) | gt == paste0(alt_allele, " ", ref_allele)) return(1) # Heterozygous = 1
    if (gt == paste0(alt_allele, " ", alt_allele)) return(2) # Homozygous (allele 2) = 2
    return(NA) # Handle unexpected genotypes
  }
  sapply(geno_column, recode) # Apply the recode function to each genotype in the column
}

# Apply the conversion function to all SNP columns in the filtered genotype data
snp_data <- filt_gen[, -1]  # Remove the "CEL file ID" column to focus on SNP markers
#num_gen_mat <- apply(snp_data, 2, convert_to_numeric)  # Apply the conversion function to each SNP column

alt_allele <- read.table("../../doc/SNP_allele_dosage", sep = "\t", header = T, row.names = 1, check.names = F)
names(alt_allele) <- "alt"

# Count occurrences of the alternative allele for each SNP marker
allele_counts <- snp_data %>%
  # For each column (SNP marker)
  imap_dfc(~ {
    # Get the alternative allele for this marker
    alt <- alt_allele[.y, "alt"]
    # Count how many times the alt allele appears in each genotype
    str_count(.x, alt)
  })

num_gen_mat <- apply(allele_counts, 2, function(x) { # Impute missing values with the mean of each column
  x[is.na(x)] <- mean(x, na.rm = TRUE)
  return(x)
})
```

# PCA

## Type

```{r}
#| label: pca-type
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center

library(FactoMineR)
pca_result <- PCA(num_gen_mat, graph = FALSE, scale.unit = TRUE)
pc_data <- data.frame(
  SampleID = filt_gen$SampleID,  # Retrieve the CEL file IDs from the original data
  PC1 = pca_result$ind$coord[, 1],  # Extract PC1 scores
  PC2 = pca_result$ind$coord[, 2],  # Extract PC2 scores
  PC3 = pca_result$ind$coord[, 3],  # Extract PC3 scores
  check.names = FALSE  # Prevents R from converting spaces to dots in column names
)
pc_data_meta <- pc_data %>% left_join(sample_info, by = "SampleID")

# Calculate the percentage of variance explained by each principal component
variance_explained <- pca_result$eig[,2]  # Second column contains % variance explained by each PC

# Create labels for the PCA plot axes, including the percentage of variance explained
pc1_label <- paste0("PC1 (", round(variance_explained[1], 2), "%)")  # Label for PC1
pc2_label <- paste0("PC2 (", round(variance_explained[2], 2), "%)")  # Label for PC2
pc3_label <- paste0("PC3 (", round(variance_explained[3], 2), "%)")  # Label for PC3

my_col <- c("#C0C0C0", "#45b6b6ff", "#51bb63ff", "#a74dafff", "#af814dff")
ggplot(pc_data_meta, aes(x = PC1, y = PC2, color = Type)) +  # Specify the data and aesthetics for the plot
  geom_point(size = 5, alpha = 0.7) +  # Add points to the plot with transparency (alpha)
  scale_color_manual(values = my_col) +  # Set custom colors for each species or population
  labs(title = "PCA - Type",  # Add a title to the plot
       x = pc1_label,  # Label the x-axis with PC1 and variance explained
       y = pc2_label) +  # Label the y-axis with PC2 and variance explained
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  geom_hline(yintercept = 0, linetype =2) +
  geom_vline(xintercept = 0, linetype =2)
```

### Permanova

```{r}
#| label: permanova-type
#| warning: false
#| message: false
#| eval: false

set.seed(100)
snp_dist <- dist(num_gen_mat)
snp_meta <- filt_gen %>% 
  left_join(sample_info, by = "SampleID") %>%
  select(!(starts_with("rs")))

adonis2(as.matrix(snp_dist) ~ Type, data = snp_meta, permutations = 999)

```

 * **Type P < 0.001**

## PASISeverity

```{r}
#| label: pca-PASISeverity
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


ggplot(pc_data_meta, aes(x = PC1, y = PC2, color = PASISeverity)) +  # Specify the data and aesthetics for the plot
  geom_point(size = 5, alpha = 0.7) +  # Add points to the plot with transparency (alpha)
  scale_color_manual(values = my_col) +  # Set custom colors for each species or population
  labs(title = "PCA - PASISeverity",  # Add a title to the plot
       x = pc1_label,  # Label the x-axis with PC1 and variance explained
       y = pc2_label) +  # Label the y-axis with PC2 and variance explained
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  geom_hline(yintercept = 0, linetype =2) +
  geom_vline(xintercept = 0, linetype =2)
```

### Permanova

```{r}
#| label: permanova-PASISeverity
#| warning: false
#| message: false
#| eval: false

set.seed(100)
adonis2(as.matrix(snp_dist) ~ PASISeverity, data = snp_meta, permutations = 999)

```

 * **PASISeverity P < 0.01**

## Gender

```{r}
#| label: pca-Gender
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


ggplot(pc_data_meta, aes(x = PC1, y = PC2, color = Gender)) +  # Specify the data and aesthetics for the plot
  geom_point(size = 5, alpha = 0.7) +  # Add points to the plot with transparency (alpha)
  scale_color_manual(values = my_col[3:4]) +  # Set custom colors for each species or population
  labs(title = "PCA - Gender",  # Add a title to the plot
       x = pc1_label,  # Label the x-axis with PC1 and variance explained
       y = pc2_label) +  # Label the y-axis with PC2 and variance explained
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  geom_hline(yintercept = 0, linetype =2) +
  geom_vline(xintercept = 0, linetype =2)
```

### Permanova

```{r}
#| label: permanova-Gender
#| warning: false
#| message: false
#| eval: false

set.seed(100)
adonis2(as.matrix(snp_dist) ~ PASISeverity, data = snp_meta, permutations = 999)

```

 * **Gender P Not Significant**

## Alcohol

```{r}
#| label: pca-Alcohol
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


ggplot(pc_data_meta, aes(x = PC1, y = PC2, color = Alcohol)) +  # Specify the data and aesthetics for the plot
  geom_point(size = 5, alpha = 0.7) +  # Add points to the plot with transparency (alpha)
  scale_color_manual(values = my_col) +  # Set custom colors for each species or population
  labs(title = "PCA - Alcohol",  # Add a title to the plot
       x = pc1_label,  # Label the x-axis with PC1 and variance explained
       y = pc2_label) +  # Label the y-axis with PC2 and variance explained
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  geom_hline(yintercept = 0, linetype =2) +
  geom_vline(xintercept = 0, linetype =2)
```

### Permanova

```{r}
#| label: permanova-Alcohol
#| warning: false
#| message: false
#| eval: false

set.seed(100)
adonis2(as.matrix(snp_dist) ~ Alcohol, data = snp_meta, permutations = 999)

```

 * **Alcohol P Not Significant**

## SmokingSnuff

```{r}
#| label: pca-SmokingSnuff
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


ggplot(pc_data_meta, aes(x = PC1, y = PC2, color = SmokingSnuff)) +  # Specify the data and aesthetics for the plot
  geom_point(size = 5, alpha = 0.7) +  # Add points to the plot with transparency (alpha)
  scale_color_manual(values = my_col) +  # Set custom colors for each species or population
  labs(title = "PCA - SmokingSnuff",  # Add a title to the plot
       x = pc1_label,  # Label the x-axis with PC1 and variance explained
       y = pc2_label) +  # Label the y-axis with PC2 and variance explained
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  geom_hline(yintercept = 0, linetype =2) +
  geom_vline(xintercept = 0, linetype =2)
```

### Permanova

```{r}
#| label: permanova-SmokingSnuff
#| warning: false
#| message: false
#| eval: false

set.seed(100)
adonis2(as.matrix(snp_dist) ~ SmokingSnuff, data = snp_meta, permutations = 999)

```

 * **SmokingSnuff P Not Significant**

```{r}
#| eval: false

num_gen_mat %>%
  as.data.frame(row.names = filt_gen$SampleID) %>%
  rownames_to_column('SampleID') %>%
  write_tsv("../snp_filtered_ordered.tsv")
```


