---
title: "Psoariasis Severity - Microbiome I"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Loading packages and running LULU

```{r}
#| label: lulu 
#| warning: false
#| message: false
#| eval: false
#| echo: false


library(lulu)
otu_table_raw <- read.table("../ampliseq_16S_output/asv_length_filter/ASV_table.len.tsv", header = T, row.names = 1)
lulu_match <- read.table("../lulu/match_list.txt")
curated_result <- lulu(otu_table_raw, lulu_match, minimum_match = 95)
lulu_otu_table <- curated_result$curated_table

write.table(lulu_otu_table, "../lulu_otu_table.tsv", sep = "\t")
```

# Filtering ASVs

## raw counts

```{r}
#| label: raw 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

library(ggplot2)
library(phyloseq)
library(dplyr)
library(tidyverse)
library(DESeq2)
library(limma)
library(ggpubr)
library(vegan)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")
sample_info <- read.table("../../data/sample_md.tsv", header = T, sep = "\t", check.names = F)
row.names(sample_info) <- sample_info$SampleID

sample_info$Type <- factor(sample_info$Type, levels = c("control", "pso_nonlesional", "pso_lesional"))
sample_info$Gender <- as.factor(sample_info$Gender)

sample_info <- sample_info %>% mutate(PsoTreatment = replace(PsoTreatment, PsoTreatment == "0", "None"))
sample_info$PsoTreatment <- factor(sample_info$PsoTreatment, levels = c("None", "Moisturizer", "TopicalSteroids"))

sample_info$PASISeverity <- as.factor(sample_info$PASISeverity)
sample_info <- sample_info %>% 
  mutate(PASISeverity = case_when(
    PASISeverity == "0" ~ "None",
    PASISeverity == "1" ~ "Low",
    PASISeverity == "2" ~ "Medium",
    PASISeverity == "3" ~ "High",
    TRUE ~ PASISeverity))
sample_info$PASISeverity <- factor(sample_info$PASISeverity, levels = c("None", "Low", "Medium", "High"))

sample_info$TypeSeverity <- factor(sample_info$TypeSeverity, 
  levels = c("control_0", "pso_nonlesional_1", "pso_nonlesional_2", "pso_nonlesional_3",
  "pso_lesional_1", "pso_lesional_2", "pso_lesional_3"))

sample_info <- sample_info %>% mutate(PrevSysTreatmentPso = replace(PrevSysTreatmentPso, PrevSysTreatmentPso == "", "None"))
sample_info$PrevSysTreatmentPso <- factor(sample_info$PrevSysTreatmentPso, levels = c("None", "No", "Yes"))

sample_info$AntiinflamMedication <- as.factor(sample_info$AntiinflamMedication)
sample_info$SSRI <- as.factor(sample_info$SSRI)
sample_info$OtherMedication <- as.factor(sample_info$OtherMedication)
sample_info$DiabetesHypertensionHyperlipidemia <- as.factor(sample_info$DiabetesHypertensionHyperlipidemia)

sample_info <- sample_info %>% 
  mutate(OtherDiseases = case_when(
    OtherDiseases == "" ~ NA,
    OtherDiseases == "0" ~ "None",
    TRUE ~ OtherDiseases))
sample_info$OtherDiseases <- as.factor(sample_info$OtherDiseases)

sample_info <- sample_info %>% mutate(SmokingSnuff = replace(SmokingSnuff, SmokingSnuff == "0", "None"))
sample_info$SmokingSnuff <- as.factor(sample_info$SmokingSnuff)

sample_info$Alcohol <- as.factor(sample_info$Alcohol)
sample_info <- sample_info %>% 
  mutate(Alcohol = case_when(
    Alcohol == "0" ~ "None",
    Alcohol == "1" ~ "Low",
    Alcohol == "2" ~ "Medium",
    Alcohol == "3" ~ "High",
    Alcohol == "4" ~ "VeryHigh",
    TRUE ~ Alcohol))
sample_info$Alcohol <- factor(sample_info$Alcohol, levels = c("None", "Low", "Medium", "High", "VeryHigh"))

sample_info <- sample_info %>% 
  mutate(Arthritis = case_when(
    Arthritis == "" ~ NA,
    Arthritis == "0" ~ "None",
    TRUE ~ Arthritis))
sample_info$Arthritis <- as.factor(sample_info$Arthritis)

sample_info <- sample_info %>% mutate(Location = replace(Location, Location == "", NA))
sample_info$Location <- as.factor(sample_info$Location)

lulu_otu_table <- read.table("../lulu_otu_table.tsv", header = T, row.names = 1, sep = "\t")

taxa_info <- read.table("../processed_taxa.tsv", header = 1, row.names = 1, sep = "\t")
taxa_info <- as.matrix(taxa_info)

OTU = otu_table(lulu_otu_table, taxa_are_rows = TRUE)
TAX = tax_table(taxa_info)
META = sample_data(sample_info)

psor_psq_raw = phyloseq(OTU, TAX, META)
psor_psq_raw
sum(otu_table(psor_psq_raw))
```

## Filtering based on taxa and counts

```{r}
#| label: filt 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

main_taxa <- c("Bacteria", "Archaea")
psor_psq_tFilt <- subset_taxa(psor_psq_raw, Kingdom %in% main_taxa)
psor_psq_tcFilt <- filter_taxa(psor_psq_tFilt, function(x) sum(x > 5) > (0.1*length(x)), TRUE)
psor_psq_filt <- subset_taxa(psor_psq_tcFilt, Family != "Mitochondria")

psor_psq_filt
sum(otu_table(psor_psq_filt))
```

More than **75%** of the counts are still kept in the dataset after the filteration steps below:
 - Only ASVs that are **Bacteria** or **Archaea** kept and the other ASVs are thrown out assuming that they are sequencing or PCR artefacts
 - ASVs that are present in atleast **10** samples with atleast **5 reads** are kept and the rest are filtered.
 - ASVs that were predicted to be **Mitochondria** are removed.

## Including --add_spe information from dada2

```{r}
#| label: addspe 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

add_spe <- read.table("../ampliseq_16S_output/dada2/ASV_tax_species.silva_138_2.tsv", header = 1, sep = "\t")
spe_exact <- add_spe %>%
  select(ASV_ID, Genus, Species_exact, confidence) %>%
  filter(Species_exact != "") %>%
  filter(confidence > 0.75) %>%
  mutate(Species_exact = paste0(Genus, '_', Species_exact)) %>%
  select(ASV_ID, Species_exact)

taxa_new <- tax_table(psor_psq_filt) %>% 
  as.data.frame() %>%
  rownames_to_column('ASV_ID') %>%
  left_join(spe_exact) %>%
  mutate(Species = ifelse(is.na(Species_exact),
    Species, Species_exact)) %>%
  select(-Species_exact)

row.names(taxa_new) <- taxa_new$ASV_ID
taxa_new$ASV_ID <- NULL
taxa_new <- as.matrix(taxa_new)
tax_table(psor_psq_filt) <- tax_table(taxa_new)

known_species <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Species)) 

known_species_asv_ids <- known_species %>%
  pull(ASV_ID)

known_genus <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Genus))

known_genus_asv_ids <- known_genus %>%
  pull(ASV_ID)

known_genus %>%
  pull(Genus) %>%
  n_distinct()

known_species %>%
  pull(Species) %>%
  n_distinct()
```

```{r}
#| eval: false
#| include: false

saveRDS(psor_psq_filt, '../psor_psq_filt.RDS')
saveRDS(known_genus_asv_ids, '../known_genus_asv_ids.RDS')
saveRDS(known_species_asv_ids, '../known_species_asv_ids.RDS')
```

There are **367 known species** and **243 known genera** found in this dataset. 


```{r}
#| label: diversity 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

psor_md <- data.frame(sample_data(psor_psq_filt))
fac_vec <- psor_md %>%
  select(where(is.factor)) %>%
  names()
num_vec <- psor_md %>%
  select(where(is.numeric)) %>%
  names()

# Alpha
alpha_phylo <- estimate_richness(psor_psq_filt, measures=c("Observed", "Shannon"))
alpha_div <- alpha_phylo %>%
  rownames_to_column("SampleID") %>%
  left_join(psor_md, by = "SampleID") %>%
  gather(Diversity_measure, value, "Shannon", "Observed")

## Beta
psor_rare <- rarefy_even_depth(psor_psq_filt, rngseed = TRUE)
psor_bray <- phyloseq::distance(psor_rare, method = "bray")
psor_mds <- phyloseq::ordinate(psor_rare, method = "MDS", distance = "bray")

pco1_var <- round(psor_mds$values$Rel_corr_eig[1] * 100, 2)
pco2_var <- round(psor_mds$values$Rel_corr_eig[2] * 100, 2)

psor_mds_12 <- psor_mds$vectors[,c(1,2)]
psor_mds_12 <- as.data.frame(psor_mds_12)
names(psor_mds_12) <- c("PCO1", "PCO2")

beta_div <- psor_mds_12 %>%
  rownames_to_column("SampleID") %>%
  left_join(psor_md, by = "SampleID") 
```

# Taxonomy Plots

## Phylum

```{r}
#| label: taxa_plot1 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

Phylum.sum = tapply(taxa_sums(psor_psq_filt), tax_table(psor_psq_filt)[, "Phylum"], sum, na.rm=FALSE)
top10Phylum = names(sort(Phylum.sum, TRUE))[1:10]
top_phylum = prune_taxa((tax_table(psor_psq_filt)[, "Phylum"] %in% top10Phylum), psor_psq_filt)

phyl_plot_list <- list()

for (a in 1:length(fac_vec)){
  phyl_plot_list[[a]] <- plot_bar(tax_glom(top_phylum, "Phylum"), "SampleID", fill="Phylum") + 
    facet_wrap(as.formula(paste("~", fac_vec[a])), scale="free_x") + 
    theme_pubclean(base_size = 20) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("#EC7063", "#A569BD", "#5DADE2", "#45B39D", "#58D68D", "#F5B041", "#DC7633", "#CACFD2", "#99A3A4", "#2E4053"))
}
names(phyl_plot_list) <- fac_vec
```

```{r}
#| label: fac_tax1
#| panel: tabset
#| output: asis
#| keep-md: true
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 14
#| fig-dpi: 300

library(purrr)
iwalk(phyl_plot_list, ~ {
  cat('## ', .y, '\n\n')
  
  print(.x)
  #print(fac_adonis_list[[.y]])
  
  cat('\n\n')
  
})

```

## Genus

```{r}
#| label: taxa_plot2
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center
#| echo: false

genus.sum = tapply(taxa_sums(psor_psq_filt), tax_table(psor_psq_filt)[, "Genus"], sum, na.rm=FALSE)
top10genus = names(sort(genus.sum, TRUE))[1:10]
top_gen = prune_taxa((tax_table(psor_psq_filt)[, "Genus"] %in% top10genus), psor_psq_filt)

gen_plot_list <- list()

for (a in 1:length(fac_vec)){
  gen_plot_list[[a]] <- plot_bar(tax_glom(top_gen, "Genus"), "SampleID", fill="Genus") + 
    facet_wrap(as.formula(paste("~", fac_vec[a])), scale="free_x") + 
    theme_pubclean(base_size = 20) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("#EC7063", "#A569BD", "#5DADE2", "#45B39D", "#58D68D", "#F5B041", "#DC7633", "#CACFD2", "#99A3A4", "#2E4053"))
}
names(gen_plot_list) <- fac_vec
```

```{r}
#| label: fac_tax2
#| panel: tabset
#| output: asis
#| keep-md: true
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 14
#| fig-dpi: 300

iwalk(gen_plot_list, ~ {
  cat('## ', .y, '\n\n')
  
  print(.x)
  #print(fac_adonis_list[[.y]])
  
  cat('\n\n')
  
})

```


# Alpha and Beta Diversity
```{r}
#| label: plot_list 
#| warning: false
#| message: false
#| echo: false

#fac_plot_list <- htmltools::tagList()
fac_plot_list <- list()
fac_adonis_list <- list()

for (a in 1:length(fac_vec)){
  # alpha-div plot
  #x <- combn(levels(eval(parse(text=paste0("psor_md$", fac_vec[a])))), 2)
  #my_comparisons <- lapply(seq_len(ncol(x)), function(i) x[,i])
  alpha_plot <- alpha_div %>%
    ggboxplot(x = fac_vec[a] ,y =  "value", fill =  fac_vec[a], facet.by = "Diversity_measure", scales = "free", add = "jitter") +
    geom_pwc(method = "t_test", label = "{p.adj.format}{p.adj.signif}", hide.ns = TRUE) +
    theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45,vjust = 0.5 )) 
  
  # beta div plot
  beta_plot <- beta_div %>%
    ggplot(aes(PCO1, PCO2)) + 
    geom_point(aes_string(color = fac_vec[a]), size = 3, show.legend = F) +
    theme_light() +
    xlab(paste0("PCO1 ", pco1_var, "%")) +
    ylab(paste0("PCO2 ", pco2_var, "%")) +
    theme(panel.grid = element_blank()) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") 
  
  #plotly
  #fac_plot_list[[a]] <- subplot(ggplotly(alpha_plot), ggplotly(beta_plot), nrows = 1)
  fac_plot_list[[a]] <- ggarrange(alpha_plot, beta_plot)

  #adonis
  if(any(is.na(eval(parse(text=paste0("psor_md$", fac_vec[a])))))){
    temp <- which(is.na(eval(parse(text=paste0("psor_md$", fac_vec[a])))))
    form <- as.formula(paste("as.matrix(psor_bray)[-temp,-temp]", fac_vec[a], sep="~"))
    fac_adonis_list[[a]]<- adonis2(form, data=psor_md[-temp,], permutations=999)
  } else {
    form <- as.formula(paste("as.matrix(psor_bray)", fac_vec[a], sep="~"))
    fac_adonis_list[[a]]<- adonis2(form, data=psor_md, permutations=999)
  }
}
names(fac_plot_list) <- fac_vec
names(fac_adonis_list) <- fac_vec
```

## Associations to metadata

```{r}
#| label: fac_tab1
#| panel: tabset
#| output: asis
#| keep-md: true
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 14
#| fig-dpi: 300

iwalk(fac_plot_list, ~ {
  cat('## ', .y, '\n\n')
  
  print(.x)
  #print(fac_adonis_list[[.y]])
  
  cat('\n\n')
  
})

```