---
title: "Psoariasis Severity - Data Integration V - Severity"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Microbiomes

The idea here is that since we have 2 microbiome for each patient with Psoriasis, we will treat these 2 microbes as 2 diffent omics from the same patient and integrate them with the SNP data. So, we will have 3 omics data that we can integrate to see how it correlates to severity.

```{r}
#| label: micr-sort
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(ggplot2)
library(phyloseq)
library(dplyr)
library(tidyverse)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")


low.count.removal <- function(
                        data, # OTU count df of size n (sample) x p (OTU)
                        percent=0.01 # cutoff chosen
                        ){
    keep.otu = which(colSums(data)*100/(sum(colSums(data))) > percent)
    data.filter = data[,keep.otu]
    return(list(data.filter = data.filter, keep.otu = keep.otu))
}

#Leisional microbiome samples
spe_glom_psq <- readRDS("../spe_glom_psq.RDS")
spe_lei <- subset_samples(spe_glom_psq, Type == "pso_lesional")
taxa_info <- as.data.frame(tax_table(spe_lei))
mic.raw <- as.data.frame(t(otu_table(spe_lei)))
names(mic.raw) <- taxa_info$Species
mic.offset <- mic.raw+1
result.filter <- low.count.removal(mic.offset, percent=0.01)
mic.lei.filter <- result.filter$data.filter
length(result.filter$keep.otu) 

#Non-Leisional microbiome samples
spe_non <- subset_samples(spe_glom_psq, Type == "pso_nonlesional")
taxa_info <- as.data.frame(tax_table(spe_non))
mic.raw <- as.data.frame(t(otu_table(spe_non)))
names(mic.raw) <- taxa_info$Species
mic.offset <- mic.raw+1
result.filter <- low.count.removal(mic.offset, percent=0.01)
mic.non.filter <- result.filter$data.filter
length(result.filter$keep.otu) 

mic.non.filter <- mic.non.filter %>%
  rownames_to_column('SampleID') %>%
  mutate(SampleID = str_replace(SampleID, "(^M.+)K$", "\\1P")) %>%
  column_to_rownames(var = "SampleID")
```

## check samples to keep

### Leisional

```{r}
#| label: micr-samp
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(mixOmics)
set.seed(100)

lib.size <- apply(mic.lei.filter, 1, sum) # determine total count for each sample
barplot(lib.size) # and plot as bar plot
mic.lei.filter <- mic.lei.filter[which(lib.size > 20000),]
mic.lei.clr <- logratio.transfo(as.matrix(mic.lei.filter), 
                             logratio = 'CLR', offset = 0)
```

The outliers `M33MK` and `M35MK` (low library size) are removed.

### Non-Leisional

```{r}
#| label: micr-samp2
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


set.seed(100)

lib.size <- apply(mic.non.filter, 1, sum) # determine total count for each sample
barplot(lib.size) # and plot as bar plot
mic.non.filter <- mic.non.filter[which(lib.size < 250000),]
mic.non.clr <- logratio.transfo(as.matrix(mic.non.filter), 
                             logratio = 'CLR', offset = 0)
```

The outliers `M16MK` (high library size) is removed.

# SNP data 

```{r}
#| label: snp
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

snp_filt <- read.table("../snp_filtered_ordered.tsv", row.names = 1, header = T)
com_samp <- Reduce(intersect, list(row.names(snp_filt), row.names(mic.lei.clr), row.names(mic.non.clr)))

snp.filt <- snp_filt[which(row.names(snp_filt) %in% com_samp), ]
snp.filt.final <- snp.filt[order(row.names(snp.filt)), ]

mic.lei.final <- mic.lei.clr[which(row.names(mic.lei.clr) %in% com_samp), ]
mic.lei.final <- mic.lei.final[order(row.names(mic.lei.final)), ] %>%
                  as.data.frame()


mic.non.final <- mic.non.clr[which(row.names(mic.non.clr) %in% com_samp), ]
mic.non.final <- mic.non.final[order(row.names(mic.non.final)), ] %>%
                  as.data.frame()

sample_info <- as.data.frame(sample_data(spe_lei))

sample_info_ord <- snp.filt %>% 
  as.data.frame() %>%
  rownames_to_column('SampleID') %>%
  dplyr::select(SampleID) %>%
  left_join(sample_info) %>% 
  column_to_rownames(var = "SampleID")
```

# Integration

```{r}
#| label: integ-I
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


library(mixOmics)
set.seed(100)

X <- list(mic_lei = mic.lei.final,
          mic_non = mic.non.final,
          snp = snp.filt.final)

Y <- factor(sample_info_ord$PASISeverity, levels = c("High", "Low", "Medium"))
summary(Y)

design <- matrix(0.1, ncol = length(X), nrow = length(X), 
                dimnames = list(names(X), names(X)))
diag(design) <- 0
design 
```

## Num. of components 

```{r}
#| label: integ-II
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

set.seed(100)
diablo.Severity <- block.plsda(X, Y, ncomp = 5, design = design)

 # For reproducibility, remove for your analyses
perf.diablo.Severity = perf(diablo.Severity, validation = 'Mfold', folds = 7, nrepeat = 10)

# Plot of the error rates based on weighted vote
plot(perf.diablo.Severity)
perf.diablo.Severity$choice.ncomp$WeightedVote
ncomp <- perf.diablo.Severity$choice.ncomp$WeightedVote["Overall.BER", "centroids.dist"]
```

## Num. of variables


```{r}
#| label: integ-III
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

set.seed(100) # for reproducibility
test.keepX <- list(mic_lei = c(5:9, seq(10, 40, 5)),
                   mic_non = c(5:9, seq(10, 40, 5)),
                   snp = c(5:9, seq(10, 20, 2)))

# tune.diablo.Severity <- tune.block.splsda(X, Y, ncomp = 2, 
#                              test.keepX = test.keepX, design = design,
#                              validation = 'Mfold', folds = 7, nrepeat = 50, 
#                              BPPARAM = BiocParallel::SnowParam(workers = 12),
#                              dist = "centroids.dist")
# list.keepX <- tune.diablo.Severity$choice.keepX

list.keepX <- list(mic_lei = c(30, 5), mic_non = c(5,6), snp = c(12, 20))

```

## Final Model

### Samples

```{r}
#| label: integ-IV
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

diablo.Severity <- block.splsda(X, Y, ncomp = 2, 
                            keepX = list.keepX, design = design)
#plotDiablo(diablo.Severity, ncomp = 1)
plotIndiv(diablo.Severity, ind.names = FALSE, legend = TRUE, 
          title = 'Severity, DIABLO comp 1 - 2', ellipse = T, abline=T)
```

```{r}
#| label: pub1
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/plotIndiv_severity.svg", width = 10, height = 7)
plotIndiv(diablo.Severity, ind.names = FALSE, legend = TRUE, 
          ellipse = T, abline=T)
dev.off()
```

### Variables

```{r}
#| label: integ-V
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
#| fig-dpi: 300
#| fig-align: center

plotVar(diablo.Severity, var.names = FALSE, style = 'graphics', legend = TRUE, 
        pch = c(16, 17, 15), cex = c(2,2,2), 
        col = c('darkorchid', 'brown1', 'lightgreen' ),
        title = 'Severity, DIABLO comp 1 - 2')

```

#### Circos Plot

```{r}
#| label: integ-VI
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
#| fig-dpi: 400
#| fig-align: center


circosPlot(diablo.Severity, cutoff = 0.7, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.variables = 1)
```

```{r}
#| label: pub2
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/circosPlot_severity.svg", width = 10, height = 10)
circosPlot(diablo.Severity, cutoff = 0.7, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.variables = 0.70, 
           size.labels = 1.5)
dev.off()
```

#### Loadings plot

```{r}
#| label: integ-VII
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

plotLoadings(diablo.Severity, comp = 1, contrib = 'max', method = 'median', size.name = 1.5)
# extracting microbiome_leisional loadings
lei_loadings_df <- data.frame(
  variable = rownames(diablo.Severity$loadings$mic_lei),
  loading = diablo.Severity$loadings$mic_lei[, 1],  # component 1
  comp = 1
)
lei_loadings_df <- lei_loadings_df %>%
                    filter(loading != 0) %>%
                    mutate(status = ifelse(loading < -0.03, "Low", ifelse(loading > 0.02, "High", "Medium")))
loadings_plot_lei <- ggplot(lei_loadings_df, aes(x = reorder(variable, abs(loading)), y = loading, fill = status)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "MIC_LEI",
    y = "Loadings value"
  ) +
  theme(axis.line.y = element_blank()) +
  scale_fill_manual(values = c("#567adf", "#df9138ff", "#aaa49eff"))

# extracting microbiome_nonleisional loadings
non_loadings_df <- data.frame(
  variable = rownames(diablo.Severity$loadings$mic_non),
  loading = diablo.Severity$loadings$mic_non[, 1],  # component 1
  comp = 1
)
non_loadings_df <- non_loadings_df %>%
                    filter(loading != 0) %>%
                    mutate(status = ifelse(loading < -0.03, "Low", ifelse(loading > 0.02, "High", "Medium")))
loadings_plot_non <- ggplot(non_loadings_df, aes(x = reorder(variable, abs(loading)), y = loading, fill = status)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "MIC_NON",
    y = "Loadings value"
  ) +
  theme(axis.line.y = element_blank()) +
  scale_fill_manual(values = c("#567adf", "#df9138ff", "#aaa49eff"))

# extracting snp loadings

snp_loadings_df <- data.frame(
  variable = rownames(diablo.Severity$loadings$snp),
  loading = diablo.Severity$loadings$snp[, 1],  # component 1
  comp = 1
)
snp_loadings_df <- snp_loadings_df %>%
                    filter(loading != 0) %>%
                    mutate(status = ifelse(loading < -0.08, "Low", ifelse(loading > 0.03, "High", "Medium")))
loadings_plot_snp <- ggplot(snp_loadings_df, aes(x = reorder(variable, abs(loading)), y = loading, fill = status)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "SNP",
    y = "Loadings value"
  ) +
  theme(axis.line.y = element_blank()) +
  scale_fill_manual(values = c("#567adf", "#df9138ff", "#aaa49eff"))
```

```{r}
#| label: pub3
#| warning: false
#| message: false
#| eval: false
library(ggpubr)
svg("../Publication_figures/loadingsPlot_severity.svg", width = 10, height = 10)
ggarrange(ggarrange(loadings_plot_lei, loadings_plot_non, ncol = 2, common.legend=T), ggarrange(loadings_plot_snp, NULL, ncol = 2), nrow = 2, common.legend=T, legend="right")
dev.off()

```

#### Heatmap

```{r}
#| label: integ-VIII
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


cimDiablo(diablo.Severity, color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
          comp = 1, margin=c(8,20), legend.position = "right")
```

```{r}
#| label: pub4
#| warning: false
#| message: false
#| eval: false

svg("../Publication_figures/heatMap_severity.svg", width = 15, height = 10)
cimDiablo(diablo.Severity, transpose = T, color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
          comp = 1, margin=c(8,30), legend.position = "topright",
          row.cex = 1.5, col.cex = 1.5, size.legend = 1)
dev.off()
```

## AUC & Prediction

```{r}
#| label: integ-IX
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center


auc.diablo.Severity.mic <- auroc(diablo.Severity, roc.block = "mic_lei", roc.comp = 2,
                   print = F)

auc.diablo.Severity.snp <- auroc(diablo.Severity, roc.block = "mic_non", roc.comp = 2,
                   print = F)

auc.diablo.Severity.snp <- auroc(diablo.Severity, roc.block = "snp", roc.comp = 2,
                   print = F)

data.test.Severity <- list(mic_lei = mic.lei.final,
          mic_non = mic.non.final,
          snp = snp.filt.final)

predict.diablo.Severity <- predict(diablo.Severity, newdata = data.test.Severity)
confusion.mat.Severity <- get.confusion_matrix(truth = droplevels(sample_info_ord$PASISeverity), 
                     predicted = predict.diablo.Severity$WeightedVote$centroids.dist[,2])
confusion.mat.Severity
```

# NMDS from latent

```{r}
#| label: nmds
#| warning: false
#| message: false
#| eval: false

cim.res <- cimDiablo(diablo.Severity, transpose = T, color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
          comp = 1, margin=c(8,30), legend.position = "topright",
          row.cex = 1.5, col.cex = 1.5, size.legend = 1)

library(vegan)
set.seed(100)
sev_dist <- dist(as.data.frame(cim.res$mat))
sev_NMDS <- metaMDS(sev_dist, k=2)

sample_info_ord$PASISeverity <- factor(sample_info_ord$PASISeverity, levels = c("High", "Low", "Medium"))

sev_NMDS_long <- sev_NMDS$points %>%
  as.data.frame() %>%
  rownames_to_column('SampleID') %>%
  left_join(sample_info_ord %>% rownames_to_column('SampleID'))

sev_nmds_plot <- ggplot(sev_NMDS_long) +
  geom_point(aes(MDS1, MDS2, color = PASISeverity), size = 4) +
  stat_ellipse(aes(MDS1, MDS2, color = PASISeverity), type = "t") +
  xlab('NMDS1') +
  ylab('NMDS2') +
  geom_hline(yintercept = 0, color = "lightgrey", linetype = 2) +
  geom_vline(xintercept = 0, color = "lightgrey", linetype = 2) +
  scale_color_manual(values = c("#567adf", "#df9138ff", "#aaa49eff")) +
  theme_light(base_size = 16) +
  theme(panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank())

svg("../Publication_figures/nmds_severity.svg", width = 10, height = 8)
sev_nmds_plot
dev.off()
```
