---
title: "Psoariasis Severity - Microbiome I"
author: "Lokesh Mano"
date: last-modified
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-location: left
    cap-location: top
    fig-cap-location: top
    code-fold: true
    code-tools: true
    mainfont: "Baskerville"
    font-size: 12
---

# Loading packages and running LULU

```{r}
#| label: lulu 
#| warning: false
#| message: false
#| eval: false


library(lulu)
otu_table_raw <- read.table("../ampliseq_16S_output/asv_length_filter/ASV_table.len.tsv", header = T, row.names = 1)
lulu_match <- read.table("../lulu/match_list.txt")
curated_result <- lulu(otu_table_raw, lulu_match, minimum_match = 95)
lulu_otu_table <- curated_result$curated_table

write.table(lulu_otu_table, "../lulu_otu_table.tsv", sep = "\t")
```

# Filtering ASVs

## raw counts

```{r}
#| label: raw 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

library(ggplot2)
library(phyloseq)
library(dplyr)
library(tidyverse)
library(DESeq2)
library(limma)
library(ggpubr)
library(vegan)
library(plotly)

theme_set(theme_bw())
set.seed(100)

setwd("/workspaces/7993_Mic_Psori/results/stats")
sample_info <- read.table("../../data/sample_md.tsv", header = T, sep = "\t", check.names = F)
row.names(sample_info) <- sample_info$SampleID

lulu_otu_table <- read.table("../lulu_otu_table.tsv", header = T, row.names = 1, sep = "\t")

taxa_info <- read.table("../processed_taxa.tsv", header = 1, row.names = 1, sep = "\t")
taxa_info <- as.matrix(taxa_info)

OTU = otu_table(lulu_otu_table, taxa_are_rows = TRUE)
TAX = tax_table(taxa_info)
META = sample_data(sample_info)

psor_psq_raw = phyloseq(OTU, TAX, META)
psor_psq_raw
sum(otu_table(psor_psq_raw))
```

## Filtering based on taxa and counts

```{r}
#| label: filt 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

main_taxa <- c("Bacteria", "Archaea")
psor_psq_tFilt <- subset_taxa(psor_psq_raw, Kingdom %in% main_taxa)
psor_psq_tcFilt <- filter_taxa(psor_psq_tFilt, function(x) sum(x > 5) > (0.1*length(x)), TRUE)
psor_psq_filt <- subset_taxa(psor_psq_tcFilt, Family != "Mitochondria")

psor_psq_filt
sum(otu_table(psor_psq_filt))
```

More than **75%** of the counts are still kept in the dataset after the filteration steps below:
 - Only ASVs that are **Bacteria** or **Archaea** kept and the other ASVs are thrown out assuming that they are sequencing or PCR artefacts
 - ASVs that are present in atleast **10** samples with atleast **5 reads** are kept and the rest are filtered.
 - ASVs that were predicted to be **Mitochondria** are removed.

## Including --add_spe information from dada2

```{r}
#| label: addspe 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

add_spe <- read.table("../ampliseq_16S_output/dada2/ASV_tax_species.silva_138_2.tsv", header = 1, sep = "\t")
spe_exact <- add_spe %>%
  select(ASV_ID, Genus, Species_exact, confidence) %>%
  filter(Species_exact != "") %>%
  filter(confidence > 0.75) %>%
  mutate(Species_exact = paste0(Genus, '_', Species_exact)) %>%
  select(ASV_ID, Species_exact)

taxa_new <- tax_table(psor_psq_filt) %>% 
  as.data.frame() %>%
  rownames_to_column('ASV_ID') %>%
  left_join(spe_exact) %>%
  mutate(Species = ifelse(is.na(Species_exact),
    Species, Species_exact)) %>%
  select(-Species_exact)

row.names(taxa_new) <- taxa_new$ASV_ID
taxa_new$ASV_ID <- NULL
taxa_new <- as.matrix(taxa_new)
tax_table(psor_psq_filt) <- tax_table(taxa_new)

known_species <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Species)) 

known_species_asv_ids <- known_species %>%
  pull(ASV_ID)

known_genus <- taxa_new %>%
  as.data.frame() %>%
  rownames_to_column("ASV_ID") %>%
  filter(!grepl("nclass", Genus))

known_genus_asv_ids <- known_genus %>%
  pull(ASV_ID)

known_genus %>%
  pull(Genus) %>%
  n_distinct()

known_species %>%
  pull(Species) %>%
  n_distinct()
```

```{r}
#| eval: false
#| include: false

saveRDS(psor_psq_filt, '../psor_psq_filt.RDS')
saveRDS(known_genus_asv_ids, '../known_genus_asv_ids.RDS')
saveRDS(known_species_asv_ids, '../known_species_asv_ids.RDS')
```

There are **367 known species** and **243 known genera** found in this dataset. 

# Alpha and Beta Diversity

```{r}
#| label: diversity 
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 20
#| fig-dpi: 400
#| fig-align: center

psor_md <- data.frame(sample_data(psor_psq_filt))

# Alpha
alpha_phylo <- estimate_richness(psor_psq_filt, measures=c("Observed", "Shannon"))
alpha_div <- alpha_phylo %>%
  rownames_to_column("SampleID")

## Beta
psor_rare <- rarefy_even_depth(psor_psq_filt, rngseed = TRUE)
psor_mds <- phyloseq::ordinate(psor_rare, method = "MDS", distance = "bray")

pco1_var <- psor_mds$values$Rel_corr_eig[1]
pco1_var <- psor_mds$values$Rel_corr_eig[2]

psor_mds_12 <- psor_mds$vectors[,c(1,2)]

```

